---
title: "NYPD_data_Rmd"
author: "Ron M"
date: "2025-06-01"
output:
  pdf_document: default
  html_document: default
---

```{r load-data, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(lubridate)
library(ggplot2)

options(width = 120)


url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
nypd_data <- read_csv(url)


nypd_data <- nypd_data %>%
  mutate(
    OCCUR_DATE = as.Date(OCCUR_DATE, format = "%m/%d/%Y"),
    Year = year(OCCUR_DATE)
  ) %>%
  filter(!is.na(OCCUR_DATE))

```{r summary}
print(summary(nypd_data), width = 100)

ggplot(nypd_data, aes(x = BORO)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Incidents by Borough",
    x = "Borough",
    y = "Number of Incidents"
  ) +
  theme_minimal()

```{r race by borough}

race_by_boro <- nypd_data %>%
  filter(!is.na(VIC_RACE), !is.na(BORO)) %>%
  group_by(BORO, VIC_RACE) %>%
  summarise(Count = n(), .groups = "drop") %>%
  arrange(desc(Count))

ggplot(race_by_boro, aes(x = BORO, y = Count, fill = VIC_RACE)) +
  geom_col(position = "dodge") +
  labs(
    title = "Victim Race by Borough",
    x = "Borough",
    y = "Number of Incidents",
    fill = "Victim Race"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10),
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  )



```{r normalized-victimization-rate, fig.width=10, fig.height=6}
# NYC Race population estimates
race_population <- tibble(
  VIC_RACE = c("BLACK", "WHITE", "ASIAN / PACIFIC ISLANDER", "HISPANIC", "AMERICAN INDIAN/ALASKAN NATIVE", "UNKNOWN"),
  Population = c(1900000, 2700000, 1350000, 2400000, 15000, 200000)
)


victim_counts <- nypd_data %>%
  filter(!is.na(VIC_RACE)) %>%
  group_by(VIC_RACE) %>%
  summarise(Count = n(), .groups = "drop")


victim_rates <- victim_counts %>%
  inner_join(race_population, by = "VIC_RACE") %>%
  mutate(Rate_per_100k = (Count / Population) * 100000)


ggplot(victim_rates, aes(x = reorder(VIC_RACE, -Rate_per_100k), y = Rate_per_100k, fill = VIC_RACE)) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "Victimization Rate per 100,000 by Race",
    x = "Victim Race",
    y = "Victimization Rate (per 100k people)"
  ) +
  theme_minimal()

yearly_counts <- nypd_data %>%
  group_by(Year) %>%
  summarise(Incidents = n()) %>%
  arrange(Year)



ggplot(yearly_counts, aes(x = Year, y = Incidents)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "steelblue") +
  scale_x_continuous(breaks = seq(min(yearly_counts$Year), max(yearly_counts$Year), 2)) +
  labs(
    title = "NYPD Incidents Over Time",
    x = "Year",
    y = "Number of Incidents"
  ) +
  theme_minimal()


borough_trends <- nypd_data %>%
  group_by(Year, BORO) %>%
  summarise(Incidents = n(), .groups = "drop")

# Plot
ggplot(borough_trends, aes(x = Year, y = Incidents, color = BORO)) +
  geom_line(size = 1) +
  geom_point() +
  scale_x_continuous(breaks = seq(min(borough_trends$Year), max(borough_trends$Year), 2)) +
  labs(
    title = "NYPD Incidents Over Time by Borough",
    x = "Year",
    y = "Number of Incidents",
    color = "Borough"
  ) +
  theme_minimal()

nypd_data <- nypd_data %>%
  mutate(Weekday = wday(OCCUR_DATE, label = TRUE))

weekday_summary <- nypd_data %>%
  group_by(Weekday) %>%
  summarise(Incidents = n())

ggplot(weekday_summary, aes(x = Weekday, y = Incidents)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Incidents by Day of Week",
    x = "Day of Week",
    y = "Number of Incidents"
  ) +
  theme_minimal()


#Bias Identification:
 # Racial biased is the most prevalent bias within this study. The data shows
#blacks are disproportionately more affected by gun violence in New YorK City than
# any other race. In order to account for this bias, I included general census
#data for New York City from 2020. 